The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-04 paper/ms.tex
Update keywords
Small changes to abstract, R0 discussion

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim60\textup{--}90\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset variation depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modelling across the large
regions of the disk accessible to {\em Gaia}.

@@

\keywords{Galaxy:~disk -- solar~neighborhood -- Galaxy:~structure --
Galaxy:~evolution}

@@

Importantly, the location of Sgr~A\textsuperscript{*} may not be equivalent to
the location of the ``dynamical Galactic center:'' the point in
three-dimensional space about which the stars in the solar neighborhood are
orbiting. This assumption, although frequently made, has not yet been
justified. It is tested in any construction of a  dynamical model where the
distance from the Sun to the Galactic center, $R_0$, is a free parameter. For
example, \citet{2015ApJ...803...80K} measured $R_0$ while modelling the
dynamics of the stream Palomar 5. Many other dynamical measurements of $R_0$
have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia}
results) but none have yet achieved a precision comparable to that of the
distance to Sgr~A\textsuperscript{*}.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim60\textup{--}90\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modelling across the large
regions of the disk accessible to {\em Gaia} and future missions.

@@

\keywords{Galaxy: disk -- Galaxy: evolution -- Galaxy: kinematics and dynamics
-- Galaxy: structure -- solar neighborhood -- stars: kinematics and dynamics}

@@

Importantly, the location of Sgr~A\textsuperscript{*} may not be equivalent to
the location of the ``dynamical Galactic center:'' the point in
three-dimensional space about which the stars in the solar neighborhood are
orbiting. This assumption, although sensible and frequently made, has not yet been
justified. It is tested in any construction of a  dynamical model where the
distance from the Sun to the Galactic center, $R_0$, is a free parameter. For
example, \citet{2015ApJ...803...80K} measured $R_0$ while modelling the
dynamics of the stream Palomar 5. Many other dynamical measurements of $R_0$
have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia}
results) but none have yet achieved a precision comparable to that of the
distance to Sgr~A\textsuperscript{*}.
